---
title: Multiple Correspondence Analysis
draft: true # true
author: Edoardo Costantini
date: '2022-05-17'
slug: mca
categories: ["PCA on categorical data"]
tags: ["PCA", "categorical"]
subtitle: ''
summary: ''
authors: ["admin"]
lastmod: '2022-04-02T14:26:57+02:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
output:
  blogdown::html_page:
    toc: true
    toc_depth: 4
    number_sections: true
---

# Introduction

Principal Component Analysis is often considered a purely descriptive tool that does not require rigorous distributional assumptions.
As such, it can be used on data that diverges from the ideal multivariate normal distribution.
It can even be used when the data is discrete.
This type of data is particularly common in the social sciences, where researchers rely on surveys for their analysis.
Surveys are often composed of a collection of items measuring the opinion or state of an individual with discrete response options.

# Learn by coding

```{r data, warning = FALSE, message = FALSE}

some <- 1

```

# TL;DR, just give me the code!
```{r TLDR, ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}
```
